---
title: "Lab 9 – Modeling Melanoma Rates - Varenya Jain"
output:
  word_document:
    toc: yes
  html_document:
    toc: yes
    df_print: paged
  html_notebook:
    theme: readable
    toc: yes
    toc_float: yes
  pdf_document: default
---

### Assignment Overview

-	Melanoma rates in different U.S. states vary—but what might explain that variation? 
-	The Melanoma.xlsx file contains melanoma rates, in addition to other potentially related variables from various sources (most of which is from 2014-2017). Note this data includes the 48 contiguous states (excludes Alaska and Hawaii).
-	We’ll also be investigating a claim made in the cancercenter.com blog linked later regarding possible risk factors for melanomas. Does their claim stack up when we look at the data?

### STEP 0

-	Pre-lab work
  - Complete the pre-lab tutorials for Lab 9 first: https://stat212-learnr.stat.illinois.edu/ 
-	Download the melanoma.xlsx file into the same folder as this RMarkdown file and then import that file into your RMarkdown file.

### Import the dataset


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(rmarkdown)
#remove.packages("rlang")
#install.packages("devtools")
#install.packages("rlang")
#install.packages("tidyverse", dependencies = TRUE)
library(tidyverse)
library(dplyr)
library(ggplot2)
tinytex::install_tinytex(force = TRUE)
cat("\014")
shell("cls")
```


```{r message=FALSE}
#library(tidyverse)
library(readxl)
Melanoma = read_excel("C:/Users/varen/Desktop/Illini+Uni/SP23/STAT_212/STAT_212-Lab_09/Melanoma-2.xlsx")
mel_data <- Melanoma
#View(mel_data)
#Your data will likely be called Melanoma.xlsx or Melanoma_1_.xlsx. Notice what name the file has when you save it to your computer.
```

### Variables

Each row of this dataset represents one patient being screened, and the following variables were documented for each patient:

- Name: State Name
- abb: State Abbreviation
- mel: Number of new melanoma cases per 10,000 people each year
- sun: The average sun exposure in each state, measured in kilojoules per square meter (kJ/m^2)
- lat: Latitude of the center point of the state 
- lon: Longitude of the center point of the state
- ocean: Whether the state borders an ocean or not
- temp: Average daily high temperature year around
- pop: State Population
- density: Average number of state residents per square miles
- age: Average age of state residents
- white: Proportion of state residents that identify as both “white” and “non-hispanic”
- income: Average income of state residents

---

## Question 1

Let’s start by creating a map of the United States and coloring each state by the Melanoma Rate. This should give us a little more insight to where melanoma rates are high.

Start by installing and librarying the package `maps`

```{r}
#install.packages("maps")
library(maps)
```

These packages contains quite a bit of map data, but we’re only going to be working on map data from U.S. states by running the following command.

```{r}
MainStates = map_data("state")
```

Next, we want to merge this data frame into our Melanoma data frame. But to do that, we need a common column. Our `Name` column in Melanoma contains state names, but in MainStates, it's called `region`. Run the following command (just copy+paste it in and run it) to change the MainStates data frame column to `Name`. Then, we will be able to easily match up these columns when we merge the data frames.

```{r}
names(MainStates)[names(MainStates) == "region"] = "Name"
```

Now we're ready to merge! Run the following function (from the dplyr package) to merge our data. Go ahead and save this merged data frame under a different name (suggested Melanoma_Map) because the dimensions are going to radically change, and it will be nice to still have the original available in this session!

```{r}
Melanoma_Map <- inner_join(MainStates, Melanoma, by = "Name")
```

We’re finally ready to **create the map! Here are your instructions:**

-	Make a ggplot with Melanoma_Map as your data
-	In the aes line, set x = long, y = lat, group = group, and fill = Mel
o	Note, these are variable names from Melanoma_Map, not from Melanoma
-	Use geom_polygon(), and set color = "black" inside this argument to insert black state borders
-	Use a fill color palette (with distiller) to choose a custom color palette. The default palette is blues, but let’s take this opportunity to pick a palette with more dynamic differences on each side of the scale!
-	Add theme_classic() to create a blank background
-	Add an appropriate title

**Which states stand out as having the highest melanoma rates?** 
**Would you say there is a geographical pattern to melanoma rates, or not?**

```{r}
#install.packages("RColorBrewer")                   # Install RColorBrewer package
library("RColorBrewer")                            # Load RColorBrewer

cat("\014")
shell("cls")
#View(Melanoma_Map)
ggplot(data = Melanoma_Map, aes(x = long, y = lat.x, group = group, fill = mel)) + labs(title = "Density of Annual Melanoma Cases in the United States", x = ": Longitude of the center point of the state", y = ": Latitude of the center point of the state") + theme_classic() +
  geom_polygon(color = "black", alpha = 0.3) +
  scale_fill_distiller(palette = "Spectral", guide = "colourbar", aesthetics = "fill") + scale_fill_fermenter(palette = "BuPu") # + scale_colour_brewer(palette = "YlOrRd", direction = -1) + scale_fill_brewer(palette = "BuPu")
#  geom_point(shape=1) + geom_smooth(method=lm) +
#  guides(fill = guide_legend(title = "", title.position = "left")) +
#  theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold"), axis.title = element_text(size = 12, face = "bold"), axis.text = element_text(size = 11, color = "hotpink4"))
print("The states which stand out as having the highest melanoma rates are Utah, Vermont, and Minnesota.")
print("From the current graph there does not appear to be any visual geographical pattern to melanoma rates.")
```





---

## Take a look at this cancer center blogpost before proceeding

Open the following short article about melanoma rates across U.S. States and read the first three paragraphs: https://www.cancercenter.com/community/blog/2016/05/unhealthiest-states-for-skin-cancer-risk-may-surprise-you 

This article is considering different possible factors that might explain why some states have higher melanoma rates. One particular factor suggested by Dr. Charles Komen Brown is that states with lower sun exposure on average tend to have higher melanoma rates. He offers one possible causal explanation by saying it is because residents in this state are not used to thinking about sun exposure and skin protection. 

---

_Be sure to use the original Melanoma data you imported from this point on. The Melanoma_Map dataframe you created was only for Question 1. _

## Question 2

Next, let's look at the relationship between a state’s average sun exposure and a state’s melanoma rate. Create a scatterplot using sun exposure as the predictor variable and melanoma rate as the response variable. However, in place of using geom_point, substitute in the following geom code:

`geom_text(aes(label=Abb),hjust=0, vjust=0, size = 3, fontface = 2)`

In addition:

-	Add a best fit line (Standard error shading optional, Color choice up to you)
-	Add a title 
-	Use the theme_classic() theme style (to provide a blank background)
-	Adjust the x and y axis titles to be fully written in 1-3 words (rather than the default variable abbreviations)

**Please include an image of this scatterplot in the report**


**Which states have higher melanoma rates—those with more sun exposure on average, or less sun exposure?**
 

**Which one or two states appear to have the largest residuals in this model?**


```{r}
cat("\014")
shell("cls")
#Local (Linear) Regression Trend via Levenberg–Marquardt algorithm Smooth Line
ggplot(data = Melanoma, aes(x = sun, y = mel)) + geom_text(aes(label = abb), hjust = 0, vjust = 0, size = 3, fontface = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "orchid") +
  ggtitle("Relationship between Sun Exposure and Melanoma Rate") +
  labs(title = "Melanoma Rate Relation to Sun Exposure per State", x = "Sun Exposure (kJ/m^2)", y = "Melanoma Rate (Annual cases per 10,000 people") +
  theme_classic()


scatterplot <- ggplot(Melanoma, aes(x=sun, y=mel)) +
  geom_text(aes(label=abb),hjust=0, vjust=0, size = 3, fontface = 2) +
  geom_smooth(method="auto") +
  ggtitle("Nonlinear Regression Relationship between Sun Exposure and Melanoma Rate") +
  xlab("Average sun exposure") +
  ylab("Melanoma rate") +
  theme_classic()
scatterplot

print("Vermont, Utah, Arizona, and Texas are unique:")
print("Utah has more sun exposure than Vermont but has approimately the same amount of cases. Arizona has the most exposure yet about half the cases. New Mexico has almost the same exposure as Arizona yet about only half the cases. Nevada has less exposure with more cases, and Texas even less exposure with the least amount of cases from all states.")


```



---

## Question 3

Now, create a simple linear model with sun exposure as a predictor of melanoma rate.

**Report the summary output of this model**


**Interpret the (multiple) r-squared value** _(using the fill in the blank template we learned in chapter 3)_

**Is there evidence that average state sun exposure and state melanoma rate have at least some linear association?** _(Don’t just say “reject” or “fail to reject” the null hypothesis—I’d recommend using the interpretational language we learned in Chapter 7)._

```{r}
cat("\014")
shell("cls")

ggplot(Melanoma, aes(x = sun, y = mel)) +
  geom_text(aes(label=abb),hjust=0, vjust=0, size = 3, fontface = 2) +
  labs(title = "Melanoma Rate Relation to Sun Exposure per State", x = "Sun Exposure (kJ/m^2)", y = "Melanoma Rate (Annual cases per 10,000 people") +
  ggtitle("Relationship between Sun Exposure and Melanoma Rate") +
  geom_smooth(method = "lm", color = "indianred2") +
  theme_classic()

mel_model <- lm(mel ~ sun, data = Melanoma)
summary(mel_model)

print("6.85% of the variance in Melanoma cases is explained by knowing the relevant Sun exposure. The p-value is also greater than 0.0723, which means we failt to reject the null hypothesis that Sun Exposure has no strong linear relationship with Melanoma Cases")


```


---

## Question 4

Let’s look at some other possible predictors of melanoma rate: Average temperature, Average age, and Proportion of white, non-hispanic residents. Create a simple linear model for _each_ of these numeric predictors (i.e., create three simple linear models) and look at the output.

**Are any of these three variables stronger predictors of melanoma rate than sun exposure? Explain. (there are at least two good ways you could decide this, but provide at least one).**

**Create a scatterplot of the strongest predictor from these three. Again, use state abbreviations in place of points. Follow all formatting guidelines from question 1. Include this plot in your report.**

**Briefly interpret the direction of this relationship and the strength of this relationship**

```{r}
cat("\014")
shell("cls")
TEMP_mel_model <- lm(mel ~ temp, data = Melanoma)
summary(TEMP_mel_model)

AGE_mel_model <- lm(mel ~ age, data = Melanoma)
summary(AGE_mel_model)

WHITE_mel_model <- lm(mel ~ white, data = Melanoma)
summary(WHITE_mel_model)


#ggplot(Melanoma, aes(x = temp, y = mel)) +
#  geom_text(aes(label=abb),hjust=0, vjust=0, size = 3, fontface = 2) +
#  labs(title = "Melanoma Rate Relation to Sun Exposure per State", x = "Average temperature", y = "Melanoma Rate (Annual cases per 10,000 people") +
#  ggtitle("Melanoma Correlation") +
#  geom_smooth(method = "lm", color = "blue") +
#  theme_classic()

#ggplot(Melanoma, aes(x = age, y = mel)) +
#  geom_text(aes(label=abb),hjust=0, vjust=0, size = 3, fontface = 2) +
#  labs(title = "Melanoma Rate Relation to Sun Exposure per State", x = "Average age", y = "Melanoma Rate (Annual cases per 10,000 people") +
#  ggtitle("Melanoma Correlation") +
#  geom_smooth(method = "lm", color = "green") +
#  theme_classic()

ggplot(Melanoma, aes(x = white, y = mel)) +
  geom_text(aes(label=abb),hjust=0, vjust=0, size = 3, fontface = 2) +
  labs(title = "Melanoma Rate Relation to Sun Exposure per State", x = "Proportion of state residents that identify as both “white” and “non-hispanic”", y = "Melanoma Rate (Annual cases per 10,000 people") +
  ggtitle("Melanoma Correlation") +
  geom_smooth(method = "lm", color = "purple") +
  theme_classic()

print("The 'white' variable has the highest r-squared and lowest p-values, 0.277 and 0.000122, with the lowest Residual standard error of 4.699 on 46 degrees of freedom")
print("There is a strong and positive linear relationship between White Identifying Residents and Melanoma cases")


```



---

## Question 5

While it is possible to create a model with two numeric predictors, let’s focus on the simpler case of using one numeric predictor with one binary predictor. 

Using an ifelse statement, create a new variable in the dataset called white_bin that records “High” if the state’s proportion of residents being “white, non-hispanic” is above 0.7 and “Low” if the proportion is below 0.7.

**Write your ifelse statement code**

**Create a scatterplot with Sun exposure on the x axis, color the data points based on whether the proportion of white residents is above 0.7 or not, and put Melanoma rate on the y axis.**

-	Continue to use the state abbreviations in place of points.
-	Add best fit lines for each group (i.e., visualize what an interaction model would look like).
-	Add an appropriate title
-	Adjust the axes labels. Also adjust the color legend to say “Proportion White.” You can do that by entering a color = “…” argument into the labs function.

```{r}
cat("\014")
library(dplyr)
shell("cls")

Melanoma <- Melanoma %>% 
  mutate(white_bin = ifelse(white > 0.7, "High", "Low"))
#Melanoma <- Melanoma %>% 
#  mutate(white_race_bin = ifelse(white > 0.7, "High", ifelse(white > 0.2, "Medium", "Low")))

ggplot(Melanoma, aes(x = white, y = mel, fill = white_bin)) +
  geom_text(aes(label=abb),hjust=0, vjust=0, size = 3, fontface = 2) +
  labs(title = "Melanoma Rate Relation to Sun Exposure per State", x = "White Resident Concentration", y = "Melanoma Rate (Annual cases per 10,000 people", color = "White Proportion") +
  ggtitle("Melanoma Correlation") +
  geom_smooth(method = "lm", color = "bisque") +
  theme_classic()

```


---

## Question 6

In the previous question, we plotted an interaction model (allowing for different slopes). Now, create an linear model with these same 2 predictors (the sun exposure variable, and white_bin) that includes an interaction term.

**Include the interaction model summary here.**

**Does the relationship between a state’s average sun exposure and that state’s melanoma rate appear to be dependent on the proportion of white, non-hispanic residents? Justify your answer using your model summary.** _It also may help you to consult your graph from the previous question to see how this visual relates to the model output._

```{r}
cat("\014")
shell("cls")
#To create a linear model with these same 2 predictors (the sun exposure variable, and white_bin) that includes an interaction term, we can use the lm() function in R. 
#The output of summary(fit) will show the coefficients of the model, including the interaction term. The interaction term represents the effect of one predictor on the response variable, depending on the level of the other predictor.
#In this case, it represents the effect of 'sun' on 'mel', depending on whether the proportion of white residents is above 0.7 or not.

fit <- lm(mel ~ sun * white_bin, data = Melanoma)
summary(fit)

print("Give the p-value of 0.04258, Multiple R-squared value of 0.1678, and Residual standard error of 5.155 on 44 degrees of freedom, it appears that a state’s average sun exposure and that state’s melanoma rate appear to be dependent on the proportion of white, non-hispanic residents")

```


---

## Question 7

Now, create an additive model for Sun exposure and white_bin as predictors of Melanoma rate.

**Include the summary of that additive model**

**Is there evidence that sun exposure has a linear relationship with melanoma rate if we already know whether the proportion white, non-hispanic residents in a state is above or below 0.7? Briefly justify**

**Contextually, how might the proportion of white, non-hispanic residents act as a confounder to the association between sun exposure and melanoma rate? (Check back to Chapter 2 and look for the confounding diagram. Think about what that might look like here!).**

```{r}
cat("\014")
shell("cls")
# plot scatter plot with color by white_bin
  #plot(df$sun, df$mel, col = ifelse(df$white_bin == "High", "red", "black"), xlab = "Sun exposure", ylab = "Melanoma rate")
ggplot(data = Melanoma, aes(x = sun, y = mel, color = white_bin)) + geom_point() + geom_text(aes(label=abb),hjust=0, vjust=0, size = 3, fontface = 2) +
  labs(title = "Melanoma Rate Relation", x = "Sun Exposure", y = "Melanoma Rate (Annual cases per 10,000 people", color = "White Proportion") +
  ggtitle("Melanoma Correlation") +
  geom_smooth(method = "lm", color = "tan2") +
  theme_classic()

#To create a linear model with these same 2 predictors (the sun exposure variable and white_bin) that includes an interaction term. We can use the lm() function to create this model:
# create the interaction model
interaction_model <- lm(mel ~ sun * white_bin, data = Melanoma)

#Next step is to create an additive model for Sun exposure and white_bin as predictors of Melanoma rate.
#Use the lm() function to create this model as well:
additive_model <- lm(mel ~ sun + white_bin + sun*white_bin, data = Melanoma)

# include the summary of the additive model
summary(additive_model)

#From notes:
  #To determine whether sun exposure has a linear relationship with melanoma rate if we already know whether the proportion white, non-hispanic residents in a state is above or below 0.7, we can look at the interaction term in the interaction model.
  #If the interaction term is significant, then there is evidence that the relationship between sun exposure and melanoma rate is different depending on the proportion of white residents.
  #If the interaction term is not significant, then there is no evidence of a different relationship between sun exposure and melanoma rate based on the proportion of white residents.

  #If the proportion of white, non-hispanic residents acts as a confounder to the association between sun exposure and melanoma rate, it means that the relationship between these two variables is being influenced by the proportion of white residents.
  #This could happen if the proportion of white residents is associated with both sun exposure and melanoma rate, and is not controlled for in the analysis.
  #In this case, the observed association between sun exposure and melanoma rate could be due to the confounding effect of the proportion of white residents, rather than a true causal relationship between sun exposure and melanoma rate.

print("We have a residual standard error of 5.155 on 44 degrees of freedom, a Multiple R-squared of 0.1678 and a p-value of 0.04258. There appears to be a strong linear relationship between melanoma rate and white proportion.")

print("Another way to consider if there is evidence that sun exposure has a linear relationship with melanoma rate if we already know whether the proportion white, non-hispanic residents in a state is above or below 0.7 is by looking at more of the additive model data. In the additive model, we can look at the coefficients for 'sun' and 'white_bin' to determine their effect on 'mel'. If the coefficient for 'sun' is significant and has the same sign in both the full model and the model with 'white_bin' included, then we can conclude that there is evidence of a linear relationship between 'sun' and 'mel', regardless of the proportion of white residents.")

print("Next is consideration of how the proportion of white, non-hispanic residents might act as a confounder to the association between sun exposure and melanoma rate. Confounding occurs when a third variable is related to both the exposure and the outcome, and it influences the observed association between the two. In this case, 'white_bin' is related to both 'sun' and 'mel', as areas with higher sun exposure and higher proportions of white residents may both have higher rates of melanoma. As such, the effect of 'sun' on 'mel' may be confounded by the proportion of white residents.")

print("An online search says that we can visualize this confounding using a something called a 'directed acyclic graph' (DAG), which shows the relationships between variables and helps identify potential confounders.")

```





## Bonus Opportunity

See the bonus questionnaire in the Chapter 14 module. It relates to making another map from this data!


